---
title: "RWorkShop_Shiny"
runtime: shiny
output: html_document
---

## Shiny Package

Shiny allows us to bring interactivity to our application. It can be useful for graphs, simulations or to be honest, almost everything where you will need interactivity. 
In Shiny there are two main function, UI (User interface) which is how the app will be displayed and the server which is the black box, where we create our functions. 

```{r}
library(tidyverse)
library(shiny)
```


```{r}

      ui <- fluidPage(
          
          titlePanel("Hello Shiny!"),

            sidebarPanel(

              sliderInput(inputId = "bins",
                          label = "Number of bins:",
                          min = 1,
                          max = 50,
                          value = 30)   
              ),

            mainPanel(
          
              plotOutput(outputId = "distPlot")
            )
        
      )
      
      
      server <- function(input, output) {

        output$distPlot <- renderPlot({

            x    <- faithful$waiting
            bins <- seq(min(x), max(x), length.out = input$bins + 1)

              hist(x, breaks = bins, col = "#75AADB", border = "white",
                   xlab = "Waiting time to next eruption (in mins)",
                   main = "Histogram of waiting times")
        })

      }
      
shinyApp(ui, server)
```    


The fluidPage function, here is a basic layout of a shiny page and it's divided in three parts:

* **titlePanel:** this is the title of the app... 
* **sidebarPanel:** the left part of the page and generally of the inputs we want 
* **mainPanel:** usually the ouputs destination

Given by the CheatSheet, we can see there are many inputs function we can use, such as a *numerics*, *sliders*, *buttons* etc.. 

Once we have defined some inputs, we will define the differents outputs we want, *plots*, *tables*, *images*... We will attribute them an *ID*, which is working like a macro variable in SAS and from that, we will get the desired value of the inputs. Here it's *bins* which will return the numeric value from the *slider*.

Then, we will create the differents outputs we would like to display on the page, these are defined with "Output" at the end of the function (*plotOuputs*, *tableOutput*.. see the Cheatsheet). As similar as the inputs, they will have an *ID* which is working in pairs with the server function. Here it's *distPlot*, We are telling, in the page we will display a plot, called *distPlot*


The server function is the "black box", this is where every computations are running. In the above exemple, we define *output$distPlot <- renderPlot*. 

So, the *renderPlot* function is working with the *plotOuput* function from the UI part. As you can see, we are calling the *plotOutput* function and we attribute it a plot.

Inside the *renderPlot* function there are all computations we need to create that plot, here it's a histogram created with the *hist()* function. 

In "bins", we are calling *input$bins* which will return the numeric value from the UI sliderInput. Each time we will change the sliderInput, bins will change as the plot will. 

**Now it's your shiny turn :)**


*** 
```{r, echo= F,message = F}

library(haven)
adsl <- read_sas("./data/adsl.sas7bdat")
adsl1 <- select(adsl, SUBJID, AGE, SEX, SMWDBL, WHOFCBL, HEIGHTBL, WEIGHTBL, TRTSDT, TRTEDT)
adsl4 <- mutate(adsl1, DaysBet = TRTEDT - TRTSDT, BMIBL = WEIGHTBL / (HEIGHTBL * HEIGHTBL * 10e-5))
adsl_ggplot <- adsl4 %>% 
  filter(!is.na(SMWDBL) & !is.na(BMIBL) & !is.na(SEX) & WHOFCBL != "")

```


1. Single quantitatives variables 
  + We will start with adsl_ggplot dataset 
  + Create adsl_shiny where there are only numeric variables
```{r}
adsl_shiny <- adsl_ggplot %>% 
  select_if(is.numeric)
```
  + Create an histogram of SMWDBL using ggplot 
```{r}
ggplot(data = adsl_ggplot, aes(x = SMWDBL)) + 
 geom_histogram()
```
  
  + Create selectInput function: 
    * "var" as inputId
    * "Choose your variable" as label
    * return the column names of adsl_shiny as choices
```{r}
selectInput(inputId = "var", label = "Choose your variable", 
            choices = names(adsl_shiny))
```
    
  + Create a renderPlot function: 
    * use adsl_ggplot as data
    * replace aes by aes_string and map x to input$var
    * add a geom_histogram, with colour = "blue", fill = "white" and y = ..density..
    * add a geom_density with alpha = 0.2 and fill = "green"
```{r}
renderPlot(
  ggplot(data = adsl_ggplot, aes_string(x = input$var)) + 
    geom_histogram(aes(y = ..density..), colour = "blue", fill = "white") + 
    geom_density(alpha = 0.2, fill = "green")
  )
```
    
  + Create a renderTable function: 
    * use summary function on adsl_shiny[,input$var]
```{r}
renderTable(
  summary(adsl_shiny[,input$var])
)
```

2. Combination of Quantitatives and Qualitatives variables  
  + Create adsl_shiny_2 from adsl_ggplot where you select columns (2:7 and 11)
```{r}
adsl_shiny_2 <- adsl_ggplot %>% 
  select(2:7,11)
```
  
  + Create selectInput function: 
    * "Yvar" as inputId
    * "Choose your Y variable" as label
    * return choice as names of all numeric variables
```{r}
selectInput(inputId = "Yvar", label = "Choose your Y variable", 
            choices = names(adsl_shiny_2 %>% select_if(is.numeric)))
```
    
  + Create selectInput function: 
    * "Xvar" as inputId
    * "Choose your X variable" as label
    * return choice as names of all character variables
```{r}
selectInput(inputId = "Xvar", label = "Choose your X variable", 
            choices = names(adsl_shiny_2 %>% select_if(is.character) ))
```
    
  + Create a renderPlot function: 
    * use adsl_ggplot as data
    * replace aes by aes_string and map x and y
    * add a geom_boxplot
    * add a geom_jitter with, width = 0.2, alpha = 0.3 and col = "red"


```{r}
renderPlot(
  ggplot(data = adsl_ggplot, aes_string(x = input$Xvar, y = input$Yvar)) + 
    geom_boxplot()+ 
    geom_jitter(width = 0.2, alpha = 0.3, col = "red")
     
  )

```

3. Combination of 2 Quantitatives and Qualitatives variables
  + Load the library "broom"
```{r}
library(broom)
```
  
  + Create adsl_shiny_2 from adsl_ggplot where you select columns (2:7 and 11)
```{r}
adsl_shiny_2 <- adsl_ggplot %>% 
  select(2:7,11)
```
  
  + Create selectInput function: 
    * "Yvar_1" as inputId
    * "Choose your Y variable" as label
    * return choice as names of all numeric variables
```{r}
selectInput(inputId = "Yvar_1", label = "Choose your Y variable", 
            choices = names(adsl_shiny_2 %>% select_if(is.numeric)))
```
    
  + Create selectInput function: 
    * "Xvar_1" as inputId
    * "Choose your X variable" as label
    * return choice as names of all numeric variables
```{r}
selectInput(inputId = "Xvar_1", label = "Choose your X variable", 
            choices = names(adsl_shiny_2 %>% select_if(is.numeric)))
```
    
  + Create selectInput function: 
    * "grp" as inputId
    * "Choose your group variable" as label
    * return choice as names of all character variables
```{r}
selectInput(inputId = "grp", label = "Choose your grouping variable", 
            choices = names(adsl_shiny_2 %>% select_if(is.character)))

```
    
  + Create a renderPlot function: 
    * use adsl_ggplot as data
    * replace aes by aes_string and map x, y and col
    * add a geom_point
    * add a smooth with method = "lm"
```{r}
renderPlot(
  ggplot(data = adsl_ggplot, aes_string(x = input$Xvar_1, y = input$Yvar_1, col = input$grp)) + 
    geom_point() + 
    geom_smooth(method = "lm")
  )
```

  + Create a renderTable function: 
    * assign formulass by filling in the following code _as.formula(paste(..., "~", ...))_
    * Here, we want to create lm_shiny  using dplyr where we will create a summary table each of the linear model, to do it, please complete the following code. 
    * call lm_shiny
  
```{r}
renderTable({
  formulass <- as.formula(paste(input$Yvar_1, "~", input$Xvar_1))

  lm_shiny <- adsl_shiny_2 %>% 
  group_by_(input$grp) %>% 
  do(tidy(lm(formulass, data=.)))
  lm_shiny
})

```

