---
title: "RWorkShop_Shiny"
runtime: shiny
output: html_document
---

## Shiny Package

Shiny allows us to bring interactivity to our application. It can be useful for graphs, simulations or to be honest, almost everything where you will need interactivity. 
In Shiny there are two main function, UI (User interface) which is how the app will be displayed and the server which is the black box, where we create our functions. 

```{r}
library( shiny )
      ui <- fluidPage(
          
          titlePanel("Hello Shiny!"),

            sidebarPanel(

              sliderInput(inputId = "bins",
                          label = "Number of bins:",
                          min = 1,
                          max = 50,
                          value = 30)   
              ),

            mainPanel(
          
              plotOutput(outputId = "distPlot")
            )
        
      )
      
      
      server <- function(input, output) {

        output$distPlot <- renderPlot({

            x    <- faithful$waiting
            bins <- seq(min(x), max(x), length.out = input$bins + 1)

              hist(x, breaks = bins, col = "#75AADB", border = "white",
                   xlab = "Waiting time to next eruption (in mins)",
                   main = "Histogram of waiting times")
        })

      }
      
shinyApp(ui, server)
```    


The fuidPage function, here is a basic layout of a shiny page and it's divided in three parts:

* **titlePanel:** this is the title of the app... 
* **sidebarPanel:** the left part of the page and generally of the inputs we want 
* **mainPanel:** usually the ouputs destination

There are many input functions to use (see cheatsheet), such as a *numerics*, *sliders*, *buttons* etc..

             sliderInput(inputId = "bins",
                          label = "Number of bins:",
                          min = 1,
                          max = 50,
                          value = 30)
            
Once we have defined some inputs, we will define the differents outputs we want, *plots*, *tables*, *images*... We will attribute them an *ID*, which is working like a macro variable in SAS and from that, we will get the desired value of the inputs. Here it's *bins* which will return the numeric value from the *slider*.


              plotOutput(outputId = "distPlot")
              
Then, we will create the differents outputs we would like to display on the page, these are defined with "Output" at the end of the function (*plotOuputs*, *tableOutput*.. see the Cheatsheet). Similar to the inputs, they will have an *ID* which works in pairs with the server function. In this example the ID is *distPlot*. We instruct the program to display a plot in the html page.

The server function contains details on the calculations, this is where every calculations are running. In the above exemple, we define *output$distPlot <- renderPlot*. 

               output$distPlot <- renderPlot({

                    x    <- faithful$waiting
                    bins <- seq(min(x), max(x), length.out = input$bins + 1)

                    hist(x, breaks = bins, col = "#75AADB", border = "white",
                         xlab = "Waiting time to next eruption (in mins)",
                         main = "Histogram of waiting times")
                })

The *renderPlot* function in the server function is the counter part to the *plotOutput* function in the UI part. In the UI we are calling the *plotOutput* function and we attribute it a plot. calculation of the plot is described in the server function.
Here it is a histogram created with the *hist()* function.

_input$bins_ will return the numeric value from the UI sliderInput.


The function seq creates a sequence vector of equidistant steps from the minimum to the maximum of the variable x (waiting time). The resulting vector has length input$bins + 1.
The function hist creates a histogram of the variable x with breaks (cut-points for the bins) as per variable bins and each time we change the sliderInput, *bins* will change as the plot will. 

**Now it's your shiny turn :)**

*** 

### Dataset

  + Import the adsl dataset 
  + Create __adsl1__ where you __select__ SUBJID, AGE, SEX, SMWDBL, WHOFCBL, HEIGHTBL, WEIGHTBL, TRTSDT, TRTEDT from adsl
  + Create __adsl4__ where you define __TrtDuration__, the number of days between TRTEDT and TRTSDT and __BMIBL__, BMI of each subject at baseline from adsl1
  + Create __adsl_ggplot__ where there is __any missing values__ in SMWDBL, BMIBL, SEX, WHOFCBL
  
```{r, message = F}
library(tidyverse)
library(shiny)
library(haven)

adsl <- read_sas(".\\data\\adsl.sas7bdat")
adsl1 <- select(adsl, SUBJID, AGE, SEX, SMWDBL, WHOFCBL, HEIGHTBL, WEIGHTBL, TRTSDT, TRTEDT)
adsl4 <- mutate(adsl1, DaysBet = TRTEDT - TRTSDT, BMIBL = WEIGHTBL / (HEIGHTBL * HEIGHTBL * 10e-5))
adsl_ggplot <- adsl4 %>% 
  filter(!is.na(SMWDBL) & !is.na(BMIBL) & !is.na(SEX) & WHOFCBL != "")

```

names(adsl)
### 1. Single quantitative variable

####- Set up the dataset

  + We will start with __adsl_ggplot__ dataset 
  + Create __adsl_shiny__ where there are only numeric variables
```{r}
adsl_shiny <- adsl_ggplot %>% 
  select_if(is.numeric)
```

  + Create an __histogram__ of __SMWDBL__ using __ggplot__ 
```{r}
ggplot(data = adsl_ggplot, aes(x = SMWDBL)) + 
 geom_histogram()
```

####- Set up the user interface function as in the example above
  + Create __selectInput__ function: 
    * "var" as __inputId__
    * "Choose your variable" as __label__
    * Column names of adsl_shiny as __choices__
    
```{r}
selectInput(inputId = "var", label = "Choose your variable", 
            choices = names(adsl_shiny))
```


*** Run the document and see if a Selectinput has been created ***


####- Set up the server function as in the example above

  + Create a __renderPlot__ function: 
    * Inside the renderPlot function, define a __ggplot__ function
    * Use __adsl_ggplot__ as __data__
    * Replace __aes__ by __aes_string__ and map x to __input$var__
    * Add a __geom_histogram__, with colour = "blue", fill = "white" and y = ..density..
    * Add a __geom_density__ with alpha = 0.2 and fill = "green"

```{r}
renderPlot(
  ggplot(data = adsl_ggplot, aes_string(x = input$var)) + 
    geom_histogram(aes(y = ..density..), colour = "blue", fill = "white") + 
    geom_density(alpha = 0.2, fill = "green")
  )
```


*** Run the document and see if a plot has been added ***

  + Create a __renderTable__ function: 
    * We will use the result of __selectInput__ to create a dynamic table
    * Inside the renderTable function, create a __summary__ function on __adsl_shiny[,input$var]__
```{r}
renderTable(
  summary(adsl_shiny[,input$var])
)
```

*** Run the document and see if a table has been added ***

###2. Combination of Quantitatives and Qualitatives variables

####- Set up the dataset
  + Create __adsl_shiny_2__ from adsl_ggplot where you select __columns 2:7 and 11__
```{r}
adsl_shiny_2 <- adsl_ggplot %>% 
  select(2:7,11)
```

####- Set up the User Interface 
  + Create __selectInput__ function: 
    * "Yvar" as __inputId__
    * "Choose your Y variable" as __label__
    * Names of all numeric variables as __choices__
```{r}
selectInput(inputId = "Yvar", label = "Choose your Y variable", 
            choices = names(adsl_shiny_2 %>% select_if(is.numeric)))
```
    
  + Create __selectInput__ function: 
    * "Xvar" as __inputId__
    * "Choose your X variable" as label
    * Names of all character variables as __choices__
```{r}
selectInput(inputId = "Xvar", label = "Choose your X variable", 
            choices = names(adsl_shiny_2 %>% select_if(is.character) ))
```
 
####- Set up the Server function     
  + Create a __renderPlot__ function: 
    * Use adsl_ggplot as __data__
    * Replace __aes__ by __aes_string__ and map x to __Xvar__ and y to __Yvar__
    * Add a __geom_boxplot__
    * Add a __geom_jitter__ with, width = 0.2, alpha = 0.3 and col = "red"

```{r}
renderPlot(
  ggplot(data = adsl_ggplot, aes_string(x = input$Xvar, y = input$Yvar)) + 
    geom_boxplot()+ 
    geom_jitter(width = 0.2, alpha = 0.3, col = "red")
  )
```


*** Run the document and see if a plot has been added ***


###3. Combination of 2 Quantitatives and Qualitative variables



####- Set up the library 

  + Load the library "broom"
```{r}
library(broom)
```


####- Set up the dataset 

  + Create __adsl_shiny_2__ from adsl_ggplot where you select __columns 2:7 and 11__
```{r}
adsl_shiny_2 <- adsl_ggplot %>% 
  select(2:7,11)
```


####- Set up the User interface  

  + Create __selectInput__ function: 
    * "Yvar_1" as __inputId__
    * "Choose your Y variable" as __label__
    * Names of all numeric variables as __choices__
```{r}
selectInput(inputId = "Yvar_1", label = "Choose your Y variable", 
            choices = names(adsl_shiny_2 %>% select_if(is.numeric)))
```
    
  + Create __selectInput__ function: 
    * "Xvar_1" as __inputId__
    * "Choose your X variable" as __label__
    * Names of all numeric variables as __choices__
```{r}
selectInput(inputId = "Xvar_1", label = "Choose your X variable", 
            choices = names(adsl_shiny_2 %>% select_if(is.numeric)))
```
    
  + Create __selectInput__ function: 
    * "grp" as __inputId__
    * "Choose your group variable" as __label__
    * Names of all character variables as __choices__
```{r}
selectInput(inputId = "grp", label = "Choose your grouping variable", 
            choices = names(adsl_shiny_2 %>% select_if(is.character)))

```
 
 
####- Set up the Server function   

  + Create a __renderPlot__ function: 
    * use adsl_ggplot as __data__
    * replace __aes__ by __aes_string__ and map x to __Xvar_1__, y to __Yvar_1__ and col to __grp__
    * add a __geom_point__
    * add a __geom_smooth__ with method = "lm"
```{r}
renderPlot(
  ggplot(data = adsl_ggplot, aes_string(x = input$Xvar_1, y = input$Yvar_1, col = input$grp)) + 
    geom_point() + 
    geom_smooth(method = "lm")
  )
```



  + Create a __renderTable__ function:
    * Here, we will create a dynamic table resuming the differents linear functions.
    * Here is how we write a linear model using R: __lm(Y ~ X, data = ...)__ with Y the dependent variable and X the predictors
    * First, assign __mod__ by filling in the following code __as.formula(paste(..., "~", ...))__ according to the previous graph and the different inputs from the user interface. __...__ are the dependent variable and the predictors 
    * Using dplyr and pipe function, create __lm_shiny__ from __adsl_shiny_2__ , grouped by the choosing __characters variable__ and then fill the blanks where we call the __lm__ of __mod__ and __data = .__

```{r}
renderTable({
  mod <- as.formula(paste(input$Yvar_1, "~", input$Xvar_1))

  lm_shiny <- adsl_shiny_2 %>% 
  group_by_(input$grp) %>% 
  do(tidy(lm(mod, data=.)))
  lm_shiny
})

```

